---
title: "NISTmAb-Native digest w/ FAIMS"
author: "Josh Baeza"
date: "`r Sys.Date()`"
output: 
  powerpoint_presentation:
    reference_doc: custom_template.potx
---


```{r setup, include=FALSE, warning = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 14, 
                      fig.height = 8)

library(tidyverse)
library(rio)
library(janitor)
library(pheatmap)
library(dendsort)
library(RColorBrewer)
library(UpSetR)

theme_set(theme_bw(base_size = 18))




```



## Experimental design

* Sample
  - 0.5 mg NIST mAb
  - 0.5 ug UPS2
  - 2 ppm PRTC
* Combined two conditions
  - Native digest with 1/1000 trypsin
  - Native digest with 1/1250 trypsin 
* Run peptides using DDA
  - For each injection, change the FAIMS CV parameter by 2 units
  - CV ranged from -20 to -110




```{r data import, include = FALSE}

psm <- import("../PD24/PSMs.txt")
fo <- import("../intermediate_data/PD_PSM_Native-digest_FAIMS-OFF.csv")
prot_list <- import("../intermediate_data/FAIMS_noFAIMS_protein_list_w_annotation.csv")
annot_file <- import("../intermediate_data/annotation_file.csv")

nist_mab <- import("../intermediate_data/Skyline_NIST_mAb_Transition_results.csv")



```


```{r, include = FALSE}

psm <- psm %>% 
  clean_names() %>% 
  mutate(spectrum_file = str_remove_all(spectrum_file, ".raw"),
         pep_length = str_length(sequence)) %>% 
  filter(master_protein_accessions != "") %>% 
  rename(filename = spectrum_file)

psm <- inner_join(annot_file, psm)


fo <- fo %>% 
  filter(techrep == 1,
         trypsin == 1000,
         master_protein_accessions != "") %>% 
  mutate(pep_length = str_length(sequence)) %>% 
  select(-digest, -condition, -sample_id, -trypsin, -techrep) %>% 
  mutate(cv = 0)


psm <- bind_rows(psm, fo)


# Adding protein annotiation
psm <- psm %>% 
  mutate(master_protein_accessions = case_when(
    master_protein_descriptions == "mAb Heavy Chain" ~ "NIST_HC",
    master_protein_descriptions == "mAb Light Chain" ~ "NIST_LC",
    TRUE ~ master_protein_accessions
  )) %>% 
  right_join(prot_list, .)


psm <- psm %>% 
  mutate(faims = ifelse(cv == 0, FALSE, TRUE)) %>% 
  relocate(faims, .after = cv)

# Summarizing Unique peptides
# using max abundance
pep <- psm %>% 
  group_by(master_protein_accessions, master_protein_descriptions, group, cv, faims,
           sequence, annotated_sequence, modifications, 
           number_of_missed_cleavages, pep_length) %>% 
  summarize(abundance = max(precursor_abundance),
            n = n()) %>% 
  ungroup()

# Summarizing protein abundances
# Summary method is max peptide abundance
prot <- psm %>% 
  group_by(master_protein_accessions, master_protein_descriptions, group, cv, faims) %>% 
  summarize(abundance = max(precursor_abundance),
            n = n()) %>% 
  ungroup()

```

```{r, include = FALSE}


pep_count_off <- pep %>% 
  filter(faims == FALSE) %>%
  arrange(-abundance) %>% 
  distinct(master_protein_accessions, master_protein_descriptions, group, sequence, modifications, .keep_all = TRUE) %>% 
  group_by(cv) %>%
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(cumulative = cumsum(count))

pep_count_faims <- pep %>% 
  filter(faims == TRUE) %>%
  arrange(-abundance) %>% 
  distinct(master_protein_accessions, master_protein_descriptions, group, sequence, modifications, .keep_all = TRUE) %>% 
  group_by(cv) %>%
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(cumulative = cumsum(count))

pep_count <- bind_rows(pep_count_off, pep_count_faims)

```

```{r, include = FALSE}



prot_count_off <- prot %>% 
  filter(faims == FALSE) %>%
  arrange(-abundance) %>% 
  distinct(master_protein_accessions, master_protein_descriptions, group, .keep_all = TRUE) %>% 
  group_by(cv) %>%
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(cumulative = cumsum(count))

prot_count_faims <- prot %>% 
  filter(faims == TRUE) %>%
  arrange(-abundance) %>% 
  distinct(master_protein_accessions, master_protein_descriptions, group, .keep_all = TRUE) %>% 
  group_by(cv) %>%
  summarize(count = n()) %>% 
  ungroup() %>% 
  mutate(cumulative = cumsum(count))

prot_count <- bind_rows(prot_count_off, prot_count_faims)


```

```{r, include = FALSE}

mab <- nist_mab %>%
  clean_names() %>%
  mutate(file_name = str_remove_all(file_name, ".raw")) %>% 
  rename(filename = file_name)

mab <- mab %>% 
  group_by(filename, protein, peptide, precursor_charge) %>%
  summarize(rt = mean(retention_time),
            abundance = sum(area),
            n = n()) %>%
  ungroup() %>% 
  left_join(annot_file, .)


pep_list <- mab %>%
  select(protein, peptide) %>%
  distinct() %>%
  mutate(kr_count = str_count(peptide, pattern = "K|R"),
         krp_count = str_count(peptide, pattern = "KP|RP"),
         missed_cleavage = kr_count - krp_count - 1) %>%
  select(protein, peptide, missed_cleavage)

mab <- full_join(pep_list, mab)


```

## Total PSMs per run

```{r}

ggplot(psm) +
  geom_bar(aes(x = cv, fill = group))

```

## Unique Peptides per run

```{r}


ggplot(pep) +
  geom_bar(aes(x = cv, fill = group))

```


## Unique Proteins per run

```{r}


ggplot(prot) +
  geom_bar(aes(x = cv, fill = group))


```


## Cumulative peptide count

```{r}

ggplot(pep_count) +
  geom_bar(aes(x = cv, y = cumulative), stat = "identity") +
  labs(title = "Peptide cumulative count")

# ggplot(pep_count) +
#   geom_bar(aes(x = cv, y = count), stat = "identity")


```

## Cumulative protein count

```{r}


ggplot(prot_count) +
  geom_bar(aes(x = cv, y = cumulative), stat = "identity") +
  labs(title = "Protein cumulative count")

# ggplot(prot_count) +
#   geom_bar(aes(x = cv, y = count), stat = "identity")


```

## Peptide count - FAIMS vs No FAIMS

```{r}


pep_faims <- pep %>% 
  select(master_protein_accessions, master_protein_descriptions, group, faims,
         sequence, annotated_sequence, modifications, 
         number_of_missed_cleavages, pep_length) %>% 
  distinct()



ggplot(pep_faims) +
  geom_bar(aes(x = faims, fill = group))



```

## Protein count - FAIMS vs No FAIMS

```{r}

prot_faims <- prot %>% 
  select(master_protein_accessions, master_protein_descriptions, group, faims) %>% 
  distinct()

ggplot(prot_faims) +
  geom_bar(aes(x = faims, fill = group))

```

## NIST mAb

```{r}




# creating a n x p table of n-peptides and p-conditions
pep_matrix <- mab %>% 
  mutate(abundance = ifelse(abundance == 0, 0, log2(abundance))) %>% 
  filter(!is.na(abundance)) %>% 
  select(cv, peptide, precursor_charge, missed_cleavage, abundance) %>% 
  unite(feature, peptide, precursor_charge, missed_cleavage, sep = "_") %>% 
  spread(cv, abundance) %>% 
  column_to_rownames("feature") %>% 
  as.matrix()

# convert all the NAs to 0. 
# NOTE: without this step, the hierarchical clustering fails
pep_matrix[is.na(pep_matrix)] <- 0

# making transposed version of the data
# This data will be used for hierarchical clustering
dat.n <- t((pep_matrix))
dat.tn <- t(scale(dat.n))

# converting 0 back to NA
pep_matrix[pep_matrix == 0] <- NA


# Distance
# Calculate distance between experiments in rows
d1 <- dist(dat.n, method = "euclidean", diag = FALSE, upper = FALSE)
# Calculate distance between peptides/proteins in rows
d2 <- dist(dat.tn, method = "euclidean", diag = FALSE, upper = FALSE)

any(is.na(d2)) # <---- must be FALSE

# hclust
# CLustering distance between experiments using Ward linkage
h1 <- fastcluster::hclust(d1, method = "ward.D2", members = NULL)
# Clustering distance between peptides/proteins using ward linkage
h2 <- fastcluster::hclust(d2, method = "ward.D2", members = NULL)

# Sorting the dendrogram
h1_sort <- as.hclust(dendsort(h1))
h2_sort <- as.hclust(dendsort(h2))

#### Protein annotation
row_annot <- data.frame(feature = rownames(pep_matrix)) %>% 
  separate(feature, into = c("peptide", "charge", "missed_cleavage"), sep = "_", remove = FALSE) %>% 
  rename(mc = missed_cleavage) %>%
  select(feature, mc, charge) %>% 
  mutate(mc = as.character(mc),
         charge = as.character(charge)) %>% 
  distinct() %>% 
  column_to_rownames("feature")


# sample_names <- data.frame(sample_id = colnames(pep_matrix)) %>% 
#   separate(sample_id, into = c("digest", "trypsin", "techrep"), sep = "-", remove = FALSE, convert = FALSE) %>% 
#   mutate(trypsin = str_pad(trypsin, width = 4, side = "left", pad = 0)) %>% 
#   unite(condition, digest, trypsin, sep = "-", remove = TRUE) %>% 
#   select(sample_id, condition) %>% 
#   column_to_rownames("sample_id")

# using custom colors for labels
my_colors <- list(
  mc = c(`0` = "#e41a1c",
         `1` = "#377eb8"
  ),
  charge = c(`2` = "#4daf4a",
             `3` = "#984ea3",
             `4` = "#ff7f00",
             `5` = "#ffff33")
  # condition = c(`DD-0400` = "#a6cee3",
  #               `DD-0800` = "#1f78b4",
  #               `DD-1000` = "#b2df8a",
  #               `DD-1250` = "#33a02c",
  #               `ND-0400` = "#fb9a99",
  #               `ND-0800` = "#e31a1c",
  #               `ND-1000` = "#fdbf6f",
  #               `ND-1250` = "#ff7f00")
)


# pheatmap
pheatmap(pep_matrix,
         scale = "none",
         # cluster_cols = h1_sort,
         cluster_cols = FALSE,
         cluster_rows = h2_sort,
         show_rownames = TRUE,
         color = colorRampPalette(brewer.pal(9, "YlGnBu"))(50),
         # color = viridis(25, direction = -1, option = "D"),
         annotation_row = row_annot,
         # annotation_col = sample_names,
         annotation_colors = my_colors,
         na_col = "grey80"
)



```

## Peptide overlap

```{r}



pep_upset <- pep %>% 
  mutate(abundance = 1) %>% 
  select(sequence, modifications, master_protein_accessions, group, cv, abundance) %>% 
  unite(feature, sequence, modifications, master_protein_accessions, group, sep = "_") %>% 
  spread(cv, abundance) %>% 
  column_to_rownames("feature")

pep_upset[!is.na(pep_upset)] <- 1
pep_upset[is.na(pep_upset)] <- 0

pep_upset <- pep_upset %>% 
  rownames_to_column("feature")


upset(pep_upset, 
      nsets = 22,
      nintersects = 18,
      keep.order = TRUE,
      order.by = "freq", 
      text.scale = 2)



```

## Peptide overlap

```{r}


features <- psm %>% 
  # filter(group %in% c("NIST")) %>% 
  select(master_protein_accessions, sequence, modifications) %>% 
  distinct() %>% 
  unite(feature, master_protein_accessions, sequence, modifications, sep = "_") %>% 
  pull()

feature_table <- psm %>% 
  # filter(group %in% c("NIST")) %>% 
  mutate(abundance = 1) %>% 
  select(master_protein_accessions, sequence, modifications, cv) %>%
  unite(feature, master_protein_accessions, sequence, modifications, sep = "_") %>% 
  distinct()



cv_list <- split(feature_table$feature, feature_table$cv)

#Now we turn the list of genes into a table of expts x genes
cv_df <- stack(setNames(cv_list, nm = paste0("CV-", names(cv_list))))


# Use the cross product to count the overlaps
cv_intersect <- table(cv_df[2:1]) %*% t(table(cv_df[2:1]))


pheatmap::pheatmap(cv_intersect, 
                   cluster_rows = FALSE,
                   cluster_cols = FALSE,
                   cellwidth = 12,
                   cellheight = 12
                   )

```

## Peptide Heatmap

```{r}


# Summarizing Unique peptides
# using max abundance
# creating a n x p table of n-peptides and p-conditions
pep_matrix <- psm %>% 
  filter(!is.na(precursor_abundance)) %>% 
  group_by(master_protein_accessions, master_protein_descriptions, group, cv, faims,
           sequence, annotated_sequence, modifications, charge,
           number_of_missed_cleavages, pep_length) %>% 
  summarize(abundance = max(precursor_abundance),
            n = n()) %>% 
  ungroup() %>% 
  mutate(abundance = log2(abundance)) %>% 
  select(master_protein_accessions, group,
         sequence, modifications, charge,
         cv, abundance) %>% 
  unite(feature, master_protein_accessions, group, sequence, modifications, charge, sep = "__") %>% 
  spread(cv, abundance) %>% 
  column_to_rownames("feature") %>% 
  as.matrix()

# convert all the NAs to 0. 
# NOTE: without this step, the hierarchical clustering fails
pep_matrix[is.na(pep_matrix)] <- 0

# making transposed version of the data
# This data will be used for hierarchical clustering
dat.n <- t((pep_matrix))
dat.tn <- t(scale(dat.n))

# converting 0 back to NA
pep_matrix[pep_matrix == 0] <- NA


# Distance
# Calculate distance between experiments in rows
d1 <- dist(dat.n, method = "euclidean", diag = FALSE, upper = FALSE)
# Calculate distance between peptides/proteins in rows
d2 <- dist(dat.tn, method = "euclidean", diag = FALSE, upper = FALSE)

# any(is.na(d2)) # <---- must be FALSE

# hclust
# CLustering distance between experiments using Ward linkage
h1 <- fastcluster::hclust(d1, method = "ward.D2", members = NULL)
# Clustering distance between peptides/proteins using ward linkage
h2 <- fastcluster::hclust(d2, method = "ward.D2", members = NULL)

# Sorting the dendrogram
h1_sort <- as.hclust(dendsort(h1))
h2_sort <- as.hclust(dendsort(h2))

#### Protein annotation
row_annot <- data.frame(feature = rownames(pep_matrix)) %>% 
  separate(feature, into = c("accession","group", "peptide", "modification", "charge"), sep = "__", remove = FALSE) %>% 
  # rename(mc = missed_cleavage) %>%
  select(feature, charge) %>% 
  mutate(charge = as.character(charge)) %>% 
  distinct() %>% 
  column_to_rownames("feature")


# sample_names <- data.frame(sample_id = colnames(pep_matrix)) %>% 
#   separate(sample_id, into = c("digest", "trypsin", "techrep"), sep = "-", remove = FALSE, convert = FALSE) %>% 
#   mutate(trypsin = str_pad(trypsin, width = 4, side = "left", pad = 0)) %>% 
#   unite(condition, digest, trypsin, sep = "-", remove = TRUE) %>% 
#   select(sample_id, condition) %>% 
#   column_to_rownames("sample_id")

# using custom colors for labels
my_colors <- list(
  charge = c(`2` = "#4daf4a",
             `3` = "#984ea3",
             `4` = "#ff7f00",
             `5` = "#ffff33",
             `6` = "#a65628",
             `7` = "#f781bf") 
)


# pheatmap
pheatmap(pep_matrix,
         scale = "row",
         # cluster_cols = h1_sort,
         cluster_cols = FALSE,
         cluster_rows = h2_sort,
         show_rownames = FALSE,
         color = colorRampPalette(brewer.pal(9, "YlGnBu"))(50),
         # color = viridis(25, direction = -1, option = "D"),
         annotation_row = row_annot,
         # annotation_col = sample_names,
         annotation_colors = my_colors,
         na_col = "grey80"
)



```

